---
title: "Reproducible Research - Peer Assessment 2:Analysis of U.S Storm Data"
author: "Lily Ip"
date: "30 January 2016"
output: html_document
---
##Synopsis

The purposes of this report is to analyse and explore the U.S. National Oceanic and Atmospheric Administration's storm database from 1950 - 2011. This data will be used to understand which types of events has most significant effects to public health and economic consequences across the United States.

The results of the below analysis shows that tornado is the cause of the most harm with respect to population health, flood causes the most property damage and drought the most crop damage,

```{r load packages, echo = FALSE, warning= FALSE, message = FALSE}
require(dplyr)
require(lubridate)
require(ggplot2)
require(knitr)
require(xlsx)
require(gridExtra)

options(stringsAsFactors = FALSE)
options(scipen=10, digits = 2)
opts_chunk$set(cache= TRUE, echo = TRUE, warning= FALSE, message = FALSE)
```
##Data Processing

In this section, we will be downloading the data set and perform basic processing for analysis.

```{r data processing, echo = TRUE}

if (!"stormData.csv.bz2" %in% dir(getwd())){
    print("downloading file")
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = paste(getwd(), "/stormData.csv.bz2", sep =  ""))
}

if (!"StormData0" %in% ls()) {
    StormData0 <-  read.csv(bzfile("stormData.csv.bz2"), header = TRUE, stringsAsFactors = FALSE)
}
dim(StormData0)
head(StormData0)

StormData0$Year <- as.numeric(format(as.Date(StormData0$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
```
The data ranges from 1950 to 2011. Below is a histogram of the number of incidents reports by year as well as a summary of the data, it shows that most of the incidents recorded occurs after 2000, most likely due to the lack of good records available.

```{r histogram}
hist(StormData0$Year, breaks = 30)
summary(StormData0$Year)
```

We now reduce the data set to the only variables needed: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP and convert the property and crop damage data into numerical forms according to the meaning of units described in the code book ([Storm  Events](http://ire.org/nicar/database-library/databases/storm-events/)). Both `PROPDMGEXP` and `CROPDMGEXP` columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r reduce data}
StormData <- StormData0 %>%
    select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```
```{r convert value of damage}
StormData$PROPDMGEXP[StormData$PROPDMGEXP == 'B'] <- 9
StormData$PROPDMGEXP[StormData$PROPDMGEXP == 'M'] <- 6
StormData$PROPDMGEXP[StormData$PROPDMGEXP == 'K'] <- 3
StormData$PROPDMGEXP[StormData$PROPDMGEXP == 'H'] <- 2
StormData$PROPDMGEXP[StormData$PROPDMGEXP == ''] <- 0
StormData$PROPDMGEXP <- as.numeric(StormData$PROPDMGEXP)
StormData$PROPDMGEXP[is.na(StormData$PROPDMGEXP)] <- 0

StormData$CROPDMGEXP[StormData$CROPDMGEXP == 'B'] <- 9
StormData$CROPDMGEXP[StormData$CROPDMGEXP == 'M'] <- 6
StormData$CROPDMGEXP[StormData$CROPDMGEXP == 'K'] <- 3
StormData$CROPDMGEXP[StormData$CROPDMGEXP == 'H'] <- 2
StormData$CROPDMGEXP[StormData$CROPDMGEXP == ''] <- 0
StormData$CROPDMGEXP <- as.numeric(StormData$CROPDMGEXP)
StormData$CROPDMGEXP[is.na(StormData$CROPDMGEXP)] <- 0

StormData$PropertyDamage <- StormData$PROPDMG * 10^StormData$PROPDMGEXP
StormData$CropDamage <- StormData$CROPDMG * 10^StormData$CROPDMGEXP
```
```{r Event type cleansing}
trim <- function (x) gsub("^[:space:]+|[:space]+$", "", x)
StormData$EVTYPE <- toupper(StormData$EVTYPE)
StormData$EVTYPE <- trim(StormData$EVTYPE)
```
##Results
### Which types of events are most harmful with respect to population and health?
```{r health and population}

Population <- StormData %>%
    group_by(EVTYPE) %>%
    summarise(TotalFatalities = sum(FATALITIES),
              TotalInjuries = sum(INJURIES))

Fatalities <- Population %>%
    select(EVTYPE, TotalFatalities) %>%
    top_n(10) %>%
    arrange(-TotalFatalities)

Injuries <- Population %>%
    select(EVTYPE, TotalInjuries) %>%
    top_n(10) %>%
    arrange(-TotalInjuries)
```
Based on the analysis, below are the two sorted lists indicating the top 10 fatalities and injuries based on event type.

####Fatalities
`r kable(Fatalities)`

####Injuries
`r kable(Injuries)`

And the following is a pair of graphs of total fatalities and total injuries affected by these severe weather events. 
```{r Polulation plot}
fatalitiesPlot <- qplot(EVTYPE, data = Fatalities, weight = TotalFatalities, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1950 - 2011")
injuriesPlot <- qplot(EVTYPE, data = Injuries, weight = TotalInjuries, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1950 - 2011")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
``` 

From the above results, we can conclude that Tornado causes the most injuries and fatalities.

###Which type of events have the greatest economic consequences?
```{r Economic}
Economic <- StormData %>%
    select(EVTYPE, PropertyDamage, CropDamage) %>%
    group_by(EVTYPE) %>%
    summarise(TotalDamage = sum(PropertyDamage) + sum(CropDamage),
              TotalProperty =  sum(PropertyDamage),
              TotalCrop = sum(CropDamage)) %>%
    top_n(10, TotalDamage) %>%
    arrange(-TotalDamage)

```

Combining the total damages to property and crop, below table indicates the to 10 event types that causes greatest economic consequences and their associated property and flood damages.
`r kable(Economic)`

And the following is a pair of graphs of total fatalities and total injuries affected by these severe weather events. 

```{r Economic plot}
propertyPlot <- qplot(EVTYPE, data = Economic, weight = TotalProperty, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+ 
    xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events in\n the U.S. from 1950 - 2011")

cropPlot<- qplot(EVTYPE, data = Economic, weight = TotalCrop, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") + 
    xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events in\n the U.S. from 1950 - 2011")
grid.arrange(propertyPlot, cropPlot, ncol = 2)
```

We can conclude that flood causes the most property damages in the US with drought as the cause of the most crop damage. However, flood is the cause of the greatest total economic consequence followed by hurricane/typhoon.

